The role of aneuploidy in the evolution of cancer drug resistance

### 2023-07-25 EvolutionaryRescue1.tex
new figures

%%%%%%%%%%%

\begin{figure}
\begin{subfigure}{0.5\textwidth}
\includegraphics[width=1\textwidth]{Figures/CombinedSubplot.pdf}
\end{subfigure}
\begin{subfigure}{0.5\textwidth}
\includegraphics[width=1\textwidth]{Figures/CombinedSubplotLargePop.pdf}
\end{subfigure}
\caption{
\textbf{Evolutionary rescue probability with partially resistant or tolerant aneuploid cells.}
Rescue probability is very high when aneuploidy provides partial resistance ($\lambda_a=0.01$), in an initially small tumor (\textbf{A}, $N=10^4$) and even more so in an initially large tumor (\textbf{B}, $N=10^8$). 
When aneuploidy provides tolerance (\textbf{C}, $N=10^4$; \textbf{D}, $N=10^8$), the rescue probability is much lower. 
In both scenarios, rescue probability increase with both the wildtype growth rate (x-axis) and the aneuploidy growth rate (colors).
Markers represent simulation results with 95\% CI; solid and dashed lines for the exact formula (\cref{eq:survival_prob} in \cref{eq:rescue_prob}); dashed lines for the approximate formula (\cref{rescue_prob_approx}), demonstrating that they all agree.
Parameters: division rate $\lambda_w=\lambda_a=\lambda_m=0.14$ (so that growth rate changes due to variable death rate); mutant death rate $\mu_m=0.13$ (so that mutant growth rate $\Delta_m=0.01$); aneuploidy rate $u=10^{-2}$; mutation rate $v=10^{-7}$.
}
\label{rescue_prob_wt_growth}
\end{figure}
%%%%%%%%%%%

\begin{figure}
% small population size
% switch from tolerance to non-growing
% TODO why no switch in panel B?
% TODO more simulations for 0 > Delta_a > Delta^*_a
% TODO simulations in A?
\begin{subfigure}{0.5\textwidth}
\includegraphics[width=1\textwidth]{Figures/P_est_divergence.pdf}
\end{subfigure}
\begin{subfigure}{0.5\textwidth}
\includegraphics[width=1\textwidth]{Figures/P_est_divergenceLargePopulation.pdf}
\end{subfigure}
\caption{\textbf{Evolutionary rescue probability with tolerant or non-growing aneuploid cells.}
Rescue probability grows with the aneuploid growth rate $\Delta_a$ (x-axis), and is much higher in an initially large tumor than in a small one (\textbf{A}: $N=10^4$; \textbf{B}: $N=10^8$). 
Markers for simulation results with 95\% CI; 
solid lines for the exact formula (\cref{eq:survival_prob} in \cref{eq:rescue_prob});
dashed lines for the approximate formula (\cref{rescue_prob_approx}).
The approximation agrees with the simulation and exact solution when the initial tumor size is large (panel B).
When the tumor size is small (panel A), we switch between the approximation for tolerant and for non-growing aneuploid cells; the switch occurs at $1/T^*$. % TODO ref to eq
Parameters: $\lambda_w=\lambda_a=\lambda_m=0.14$; $\mu_m=0.13$; $u=10^{-2}$; $v=10^{-7}$.
}
\label{rescue_prob_an_growth}
\end{figure}

%%%%%%%%%%%

\begin{figure}
\begin{subfigure}{0.5\textwidth}
% non-growing slightly declining, compare 1st order approximation to exact and simulation
\includegraphics[width=1\textwidth]{Figures/DeleteriousTauLeapPlot.pdf}
\end{subfigure}
\begin{subfigure}{0.5\textwidth}
% density-dependent simulations, barely growing (or maybe partial resistance), with exact formula
\includegraphics[width=1\textwidth]{Figures/SurvPlotNDataLogisticK.pdf}
\end{subfigure}
\caption{\textbf{Evolutionary rescue probability for variable initial tumor size.}
\textbf{(A)} Comparison of simulation results (markers with 95\% CI, too small to appear with $10^5$ simulations per marker), the exact formula (blue line, \cref{eq:survival_prob} in \cref{eq:rescue_prob}) and the approximate formula (orange line, \cref{rescue_prob_approx}).
\textbf{(B)} Comparison of results of simulations  with density-dependent growth (markers with with 95\% CI) and the exact formula (blue line, \cref{eq:survival_prob} in \cref{eq:rescue_prob}) with maximum carrying capacity $K=10^9$.
Parameters: $\lambda_w=\lambda_a=\lambda_m=0.14$; $\mu_w=0.17$; (A) $\mu_a=0.15$, (B) $\mu_a=0.135$; $\mu_m=0.13$; $u=10^{-2}$; $v=10^{-7}$.
}
\label{rescue_prob_N}
\end{figure}

%%%%%%%%%%%

\begin{figure}
% slightly decreasing (need to check)
 \vspace*{1\baselineskip}
\includegraphics[width=1\textwidth]{Figures/FractionPlot.pdf}
\caption{\textbf{Effect of standing variation on evolutionary rescue.}
In aneuploid cells already exist in the population at the onset of drug therapy as standing genetic variation, then evolutionary rescue is more likely... 
Parameters: $\lambda_w=\lambda_a=\lambda_m=0.14$; $\mu_w=0.17$; $\mu_a=0.145$; $\mu_m=0.13$; $u=10^{-2}$; $v=10^{-7}$.
% TODO think about this figure...
% TODO maybe better to show p_total / p_denovo as function of u or N.
}
\label{FractionPlot}
\end{figure}

%%%%%%%%%%%

\begin{figure}
\begin{subfigure}{0.5\textwidth}
\includegraphics[width=1\textwidth]{Figures/RatioEvolRescue.pdf}
\end{subfigure}
\begin{subfigure}{0.5\textwidth}
\includegraphics[width=1\textwidth]{Figures/RatioEvolRescuePopulationSize.pdf}
\end{subfigure}
\caption{\textbf{Effect of aneuploidy on evolutionary rescue.}
The ratio of rescue probability with and without aneuploid ($H$, \cref{ratiorescue}) increases with the aneuploid growth rate (colors) and decreases with the wildtype growth rates and initial tumor size (x-axes), except for large tumors where where the ratio converges to unity.
\textbf{(A, B)} Aneuploidy provides partial resistance.
\textbf{(C, D)} Aneuploidy provides tolerance.  
Solid and dashed lines apply $p_{rescue}$ from the exact formula of  (\cref{eq:survival_prob} in \cref{eq:rescue_prob}); dashed lines apply $p_{rescue}$ from the approximate formula (\cref{rescue_prob_approx}), with good agreement.
Parameters: $N=10^4$; $\lambda_w=\lambda_a=\lambda_m=0.14$; (B) $\mu_w=0.17$; $\mu_m=0.13$; $u=10^{-2}$; $v=10^{-7}$.
% TODO start y at 1
}
\label{rescue_ratio}
\end{figure}

%%%%%%%%%%%

\begin{figure}
\begin{subfigure}{0.5\textwidth}
\includegraphics[width=1\textwidth]{Figures/MeanTimeDeleteriousPlot.pdf}
\end{subfigure}
\begin{subfigure}{0.5\textwidth}
\includegraphics[width=1\textwidth]{Figures/MeanTimeGrowthMutantDirectPlot.pdf}
\end{subfigure}
\caption{\textbf{Evolutionary rescue time.}
Shown is the mean time for appearance of a resistance mutation the leads to evolutionary rescue \textbf{(left)} with ($u>0$) and \textbf{(right)} without ($u=0$ aneuploidy.
Our inhomogeneous Poisson-process approximations (solid blue lines, right: \cref{eq:meantime1}, left: \cref{meantimet2}) is in agreement with simulation results (orange markers with 95\% CI).
Our 1st-order (dashed red lines, right: \cref{eq:limitapprox}, left: \cref{limitapprox2}) and 2nd-order (green line, left: \cref{limitapprox3}) approximations work well when the initial tumor size is large (here $>10^8$ cells).
Parameters: $\lambda_w=\lambda_a=\lambda_m=0.14$; $\mu_w=0.17$; (A) $\mu_a=0.145$; $\mu_m=0.13$; $u=10^{-2}$; $v=10^{-7}$.
}
\label{MeanTimeGrowthAneuploidyPlot} 
\end{figure}

%%%%%%%%%%%

\begin{figure}
 \vspace*{1\baselineskip}
\includegraphics[width=1\textwidth]{Figures/MeanTimeRatioInitialPopulationSize.pdf}
\caption{\textbf{Ratio of evolutionary rescue time with and without aneuploidy.}
The ratio of the mean time to appearance of a resistance mutation that leads to evolutionary rescue with ($u>0$) and without ($u=0$) aneuploidy for variable initial tumor sizes (\cref{eq:eqMeanTimeRatioInitialPopulationSize}) when aneuploidy provides tolerance to the drug ($\Delta_a < 0$). When the initial tumor size is not large ($<10^8$), aneuploidy can decrease the rescue time by 10-100-fold.
% TODO parameter values 
}
\label{MeanTimeRatioInitialPopulationSize}
\end{figure}

%%%%%%%%%%%

\begin{figure}
 \vspace*{1\baselineskip}
\includegraphics[width=1\textwidth]{Figures/ThresholdPopulationSize.pdf}
\caption{\textbf{TODO.}
TODO
}
\label{MinTumorSize}
\end{figure}

>>>>

%%%%%%%%%%%
% Fig 3A: prob rescue vs N for various Delta_a; markup N*
% Fig 3B: N* vs Delta_a
% Fig 3C: N* vs u/v

\begin{figure}
\begin{subfigure}{0.5\textwidth}
A\\
\includegraphics[width=1\textwidth]{Figures/ProbvNPlot.pdf}
\end{subfigure}
\\
\begin{subfigure}{0.5\textwidth}
B\\
\includegraphics[width=1\textwidth]{Figures/ThresholdPopulationSizePlot.pdf}
\end{subfigure}
\begin{subfigure}{0.5\textwidth}
C\\
\includegraphics[width=1\textwidth]{Figures/ThresholdPopulationSizeVersusRatioPlot.pdf}
\end{subfigure}
\caption{
\textbf{Aneuploidy facilitates evolutionary rescue of cancer under drug treatment.}
\textbf{(A)} The probability of evolutionary rescue (i.e. the probability that the population does not go to extinction), $\presc$, as a function of the initial tumor size, $N$. Dashed line shows the threshold tumor size, $N^*$, above which the probability is very high.
\textbf{(B)} The threshold tumor size $N_a^*$ as a function of the aneuploid growth rate $\Delta_a$. The dashed horizontal line shows $N^*_m$, the threshold tumor size without aneuploidy ($u=0$). When aneuploid growth rate is close to or higher than zero, aneuploidy decreases the threshold tumor size, thereby facilitating evolutionary rescue.
\textbf{(C)} The threshold tumor size $N_a^*$ as a function of the ratio of aneuploidy and mutation rates, $u/v$. The dashed horizontal line shows $N^*_m$, the threshold tumor size without aneuploidy ($u=0$).  When the aneuploidy rate is much higher than the mutation rate, aneuploidy decreases the threshold tumor size, thereby facilitating evolutionary rescue.
}
\label{fig:LABEL}
\end{figure}
%%%%%%%%%%%

%%%%%%%%%%%%
%
%\begin{figure}
%\begin{subfigure}{0.5\textwidth}
%\includegraphics[width=1\textwidth]{Figures/FILENAME.pdf}
%\end{subfigure}
%\begin{subfigure}{0.5\textwidth}
%\includegraphics[width=1\textwidth]{Figures/FILENAME.pdf}
%\end{subfigure}
%\caption{
%\textbf{TITLE}
%CAPTION
%}
%\label{fig:LABEL}
%\end{figure}
%%%%%%%%%%%%

%%%%%%%%%%%%
%
%\begin{figure}
%\begin{subfigure}{0.5\textwidth}
%\includegraphics[width=1\textwidth]{Figures/CombinedSubplot.pdf}
%\end{subfigure}
%\begin{subfigure}{0.5\textwidth}
%\includegraphics[width=1\textwidth]{Figures/CombinedSubplotLargePop.pdf}
%\end{subfigure}
%\caption{
%\textbf{Evolutionary rescue probability with partially resistant or tolerant aneuploid cells.}
%Rescue probability is very high when aneuploidy provides partial resistance ($\lambda_a=0.01$), in an initially small tumor (\textbf{A}, $N=10^4$) and even more so in an initially large tumor (\textbf{B}, $N=10^8$). 
%When aneuploidy provides tolerance (\textbf{C}, $N=10^4$; \textbf{D}, $N=10^8$), the rescue probability is much lower. 
%In both scenarios, rescue probability increase with both the wildtype growth rate (x-axis) and the aneuploidy growth rate (colors).
%Markers represent simulation results with 95\% CI; solid and dashed lines for the exact formula (\cref{eq:survival_prob} in \cref{eq:rescue_prob}); dashed lines for the approximate formula (\cref{rescue_prob_approx}), demonstrating that they all agree.
%Parameters: division rate $\lambda_w=\lambda_a=\lambda_m=0.14$ (so that growth rate changes due to variable death rate); mutant death rate $\mu_m=0.13$ (so that mutant growth rate $\Delta_m=0.01$); aneuploidy rate $u=10^{-2}$; mutation rate $v=10^{-7}$.
%}
%\label{rescue_prob_wt_growth}
%\end{figure}
%%%%%%%%%%%%
%
%\begin{figure}
%% small population size
%% switch from tolerance to non-growing
%% TODO why no switch in panel B?
%% TODO more simulations for 0 > Delta_a > Delta^*_a
%% TODO simulations in A?
%\begin{subfigure}{0.5\textwidth}
%\includegraphics[width=1\textwidth]{Figures/P_est_divergence.pdf}
%\end{subfigure}
%\begin{subfigure}{0.5\textwidth}
%\includegraphics[width=1\textwidth]{Figures/P_est_divergenceLargePopulation.pdf}
%\end{subfigure}
%\caption{\textbf{Evolutionary rescue probability with tolerant or non-growing aneuploid cells.}
%Rescue probability grows with the aneuploid growth rate $\Delta_a$ (x-axis), and is much higher in an initially large tumor than in a small one (\textbf{A}: $N=10^4$; \textbf{B}: $N=10^8$). 
%Markers for simulation results with 95\% CI; 
%solid lines for the exact formula (\cref{eq:survival_prob} in \cref{eq:rescue_prob});
%dashed lines for the approximate formula (\cref{rescue_prob_approx}).
%The approximation agrees with the simulation and exact solution when the initial tumor size is large (panel B).
%When the tumor size is small (panel A), we switch between the approximation for tolerant and for non-growing aneuploid cells; the switch occurs at $1/T^*$. % TODO ref to eq
%Parameters: $\lambda_w=\lambda_a=\lambda_m=0.14$; $\mu_m=0.13$; $u=10^{-2}$; $v=10^{-7}$.
%}
%\label{rescue_prob_an_growth}
%\end{figure}
%
%%%%%%%%%%%%
%
%\begin{figure}
%\begin{subfigure}{0.5\textwidth}
%% non-growing slightly declining, compare 1st order approximation to exact and simulation
%\includegraphics[width=1\textwidth]{Figures/DeleteriousTauLeapPlot.pdf}
%\end{subfigure}
%\begin{subfigure}{0.5\textwidth}
%% density-dependent simulations, barely growing (or maybe partial resistance), with exact formula
%\includegraphics[width=1\textwidth]{Figures/SurvPlotNDataLogisticK.pdf}
%\end{subfigure}
%\caption{\textbf{Evolutionary rescue probability for variable initial tumor size.}
%\textbf{(A)} Comparison of simulation results (markers with 95\% CI, too small to appear with $10^5$ simulations per marker), the exact formula (blue line, \cref{eq:survival_prob} in \cref{eq:rescue_prob}) and the approximate formula (orange line, \cref{rescue_prob_approx}).
%\textbf{(B)} Comparison of results of simulations  with density-dependent growth (markers with with 95\% CI) and the exact formula (blue line, \cref{eq:survival_prob} in \cref{eq:rescue_prob}) with maximum carrying capacity $K=10^9$.
%Parameters: $\lambda_w=\lambda_a=\lambda_m=0.14$; $\mu_w=0.17$; (A) $\mu_a=0.15$, (B) $\mu_a=0.135$; $\mu_m=0.13$; $u=10^{-2}$; $v=10^{-7}$.
%}
%\label{rescue_prob_N}
%\end{figure}
%
%%%%%%%%%%%%
%
%\begin{figure}
%% slightly decreasing (need to check)
% \vspace*{1\baselineskip}
%\includegraphics[width=1\textwidth]{Figures/FractionPlot.pdf}
%\caption{\textbf{Effect of standing variation on evolutionary rescue.}
%In aneuploid cells already exist in the population at the onset of drug therapy as standing genetic variation, then evolutionary rescue is more likely... 
%Parameters: $\lambda_w=\lambda_a=\lambda_m=0.14$; $\mu_w=0.17$; $\mu_a=0.145$; $\mu_m=0.13$; $u=10^{-2}$; $v=10^{-7}$.
%% TODO think about this figure...
%% TODO maybe better to show p_total / p_denovo as function of u or N.
%}
%\label{FractionPlot}
%\end{figure}
%
%%%%%%%%%%%%
%
%\begin{figure}
%\begin{subfigure}{0.5\textwidth}
%\includegraphics[width=1\textwidth]{Figures/RatioEvolRescue.pdf}
%\end{subfigure}
%\begin{subfigure}{0.5\textwidth}
%\includegraphics[width=1\textwidth]{Figures/RatioEvolRescuePopulationSize.pdf}
%\end{subfigure}
%\caption{\textbf{Effect of aneuploidy on evolutionary rescue.}
%The ratio of rescue probability with and without aneuploid ($H$, \cref{ratiorescue}) increases with the aneuploid growth rate (colors) and decreases with the wildtype growth rates and initial tumor size (x-axes), except for large tumors where where the ratio converges to unity.
%\textbf{(A, B)} Aneuploidy provides partial resistance.
%\textbf{(C, D)} Aneuploidy provides tolerance.  
%Solid and dashed lines apply $p_{rescue}$ from the exact formula of  (\cref{eq:survival_prob} in \cref{eq:rescue_prob}); dashed lines apply $p_{rescue}$ from the approximate formula (\cref{rescue_prob_approx}), with good agreement.
%Parameters: $N=10^4$; $\lambda_w=\lambda_a=\lambda_m=0.14$; (B) $\mu_w=0.17$; $\mu_m=0.13$; $u=10^{-2}$; $v=10^{-7}$.
%% TODO start y at 1
%}
%\label{rescue_ratio}
%\end{figure}
%
%%%%%%%%%%%%
%
%\begin{figure}
%\begin{subfigure}{0.5\textwidth}
%\includegraphics[width=1\textwidth]{Figures/MeanTimeDeleteriousPlot.pdf}
%\end{subfigure}
%\begin{subfigure}{0.5\textwidth}
%\includegraphics[width=1\textwidth]{Figures/MeanTimeGrowthMutantDirectPlot.pdf}
%\end{subfigure}
%\caption{\textbf{Evolutionary rescue time.}
%Shown is the mean time for appearance of a resistance mutation the leads to evolutionary rescue \textbf{(left)} with ($u>0$) and \textbf{(right)} without ($u=0$ aneuploidy.
%Our inhomogeneous Poisson-process approximations (solid blue lines, right: \cref{eq:meantime1}, left: \cref{meantimet2}) is in agreement with simulation results (orange markers with 95\% CI).
%Our 1st-order (dashed red lines, right: \cref{eq:limitapprox}, left: \cref{limitapprox2}) and 2nd-order (green line, left: \cref{limitapprox3}) approximations work well when the initial tumor size is large (here $>10^8$ cells).
%Parameters: $\lambda_w=\lambda_a=\lambda_m=0.14$; $\mu_w=0.17$; (A) $\mu_a=0.145$; $\mu_m=0.13$; $u=10^{-2}$; $v=10^{-7}$.
%}
%\label{MeanTimeGrowthAneuploidyPlot} 
%\end{figure}
%
%%%%%%%%%%%%
%
%\begin{figure}
% \vspace*{1\baselineskip}
%\includegraphics[width=1\textwidth]{Figures/MeanTimeRatioInitialPopulationSize.pdf}
%\caption{\textbf{Ratio of evolutionary rescue time with and without aneuploidy.}
%The ratio of the mean time to appearance of a resistance mutation that leads to evolutionary rescue with ($u>0$) and without ($u=0$) aneuploidy for variable initial tumor sizes (\cref{eq:eqMeanTimeRatioInitialPopulationSize}) when aneuploidy provides tolerance to the drug ($\Delta_a < 0$). When the initial tumor size is not large ($<10^8$), aneuploidy can decrease the rescue time by 10-100-fold.
%% TODO parameter values 
%}
%\label{MeanTimeRatioInitialPopulationSize}
%\end{figure}
%
%%%%%%%%%%%%
%
%\begin{figure}
% \vspace*{1\baselineskip}
%\includegraphics[width=1\textwidth]{Figures/ThresholdPopulationSize.pdf}
%\caption{\textbf{TODO.}
%TODO
%}
%\label{MinTumorSize}
%\end{figure}
